Tools for identifying unexpectedly low microfossil count sums

### 2020-06-12 Rmd/count_check_MS.Rmd
datasets pollen23 plots and tables

In datasets pollen2 and pollen3, presented in the same paper, # % and # % of assemblages do not have singletons and have a greatest common divisor greater than one.
Both datasets match the published pollen stratigraphies.
For most of both datasets, assemblages with and without singletons are arranged alternately.
Samples without singletons are on the 10 cm intervals, and those with on the intervening 5 cm intervals. 
Closer inspection reveals that the assemblages between two assemblages without singletons have counts that are the mean of their neighbours.
On request, the authors sent the original count data. 
Comparison of these with the archived data showed that the archived assemblages without singletons had had their counts multiplied by two.
The intervening assemblages did not exist in the original data. 
In the published and archived data, they had been interpolated from their neighbours (Fig. \@ref(fig:archived-original-plot-plot)).

fig - plot of 1 or 2 mid-common taxa by depth in archived and original

```{r neotoma-table}
pollen1234$table %>% knitr::kable(caption = "The number of assemblages, median number of taxa, percent of assemblages with a greatest common divisor (GCD) of one, and the median number of singletons in assemblages with a greatest common divisor of one, in four pollen datasets")
```

```{r archived-original-plot, fig.height = 70/24.5, fig.width=80/24.5, fig.cap="Number of taxa in against count sum for assemblages in two pollen datasets with a greatest common divisor of one (red circles) or two (blue triangles). Lines are loess with a span of 1. All four pollen datasets show the same pattern, but it is clearest in datasets 1 and 2 which have the highest proportion of assemblages with a greatest common divisor of two.", message = FALSE}
readd(archived_original_plot)
```

One assemblage with a greatest common divisor of two has `r pollen1234$summ$max_richness` taxa. 
Assuming the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$: for a single assemblage with this many taxa, the probability is 1 in 10^`r floor(log10(2^pollen1234$summ$max_richness))`^.
The probability that these counts reflect the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled at some stage in some way.

>>>>

In datasets pollen2 and pollen3, presented in the same paper, # % and # % of assemblages do not have singletons and have a greatest common divisor greater than one (Table \@ref(tab:unexpected-table))).
Both datasets match the published pollen stratigraphies.
For most of both datasets, assemblages with and without singletons are arranged alternately.
Samples without singletons are on the 10 cm intervals, and those with on the intervening 5 cm intervals. 
Closer inspection reveals that the assemblages between two assemblages without singletons have counts that are the mean of their neighbours.
On request, the authors sent the original count data. 
Comparison of these with the archived data showed that the archived assemblages without singletons had had their counts multiplied by two.
The intervening assemblages did not exist in the original data. 
In the published and archived data, they had been interpolated from their neighbours (Fig. \@ref(fig:interpolated-plot)).

```{r unexpected-table}
readd(unexpected_table) %>% knitr::kable(caption = "The number of assemblages, median number of taxa, percent of assemblages with a greatest common divisor (GCD) of one, and the median number of singletons in assemblages with a greatest common divisor of one, in four pollen datasets")
```

```{r interpolated-plot, warning = FALSE, fig.height = 100/24.5, fig.width=180/24.5, fig.cap="A) Plot of selected taxa by depth from dataset pollen2. Counts with blue triangles are from assemblages with a greatest common divisor of one. These appear to have been linearly interpolated between the adjacent counts from assemblages with a greatest common divisor of two (red circles). B) Percent of taxa in each assemblage where the count is the mean of the counts in the adjacent assemblages, excluding runs of zero counts. Most of the assemblages with a greatest common divisor of one are exactly linearly interpolated between adjacent assemblages. The main exception is the assemblage at 210 cm which has a single singleton and otherwise has a greatest common divisor of two, other exceptions have a few extra singletons not predicted from the adjacent assembages.", message = FALSE}
readd(pollen23_plot)
```

One assemblage with a greatest common divisor of two has `r "pollen1234$summ$max_richness"` taxa. 
Assuming the distribution of counts is (at least locally) uniform, the probability of _n_ counts being divisible by _k_ is $1/k^n$: for a single assemblage with this many taxa, the probability is 1 in 10^`r "floor(log10(2^pollen1234$summ$max_richness))"`^.
The probability that these counts reflect the true nature of the pollen assemblages is exceedingly low; it is far more likely that the data have been mishandled at some stage in some way.
